### 2022-10-09 AISTATS2023/paper.tex
update

Recently, Machine Learning (ML) has been adopted in critical domains,
such as healthcare and finance. In these areas, we need a combination
of accurate predictions along with meaningful explanations to support
them. For this reason there is an increased interest in Explainable AI
(XAI), to understand the decision mechanism of complex black-box
models. XAI literature distinguishes between local and global
techniques~\citep{Molnar2020interpretable}. Local methods provide
instance-level explanations, i.e.~explain the prediction for a
specific instance, whereas global methods summarize the entire model
behavior.

Global methods create a global explanation by aggregating the
instance-level explanations into a single interpretable outcome,
usually a number or a plot. When strong interactions between features
exist, the instance-level explanations may significantly deviate from
the aggregated outcome~\citep{herbinger2022repid}, a phenomenon called
\emph{aggregation bias}~\citep{mehrabi2021survey}. Therefore, global
methods should quantify not only the aggregated (averaged) effect, but
also the level of heterogeneity of instance-level explanations. The
same holds for Feature Effect (FE)~\citep{Gromping2020MAEP}, a class
of global XAI methods that isolate the impact of a single feature on
the output.\footnote{FE methods also isolate the combined effect of a
  pair of features to the output. Combinations of more than two
  features are uncommon, since they are difficult to estimate and
  visualize.} For example, a feature with zero average effect may
indicate no effect on the output or a highly positive effect in some
cases and a highly negative effect in others. The ambiguity about the
correct interpretation could be resolved, if the FE method informs
about the heterogeneity of instance-level explanations, a quantity we
refer to as the \emph{uncertainty of the effect}.

@@

In this section, we describe the basic methods for FE and uncertainty
quantification, focusing on ALE definition. We, then, review the ALE
approximation\citep{apley2020visualizing, gkolemis22}, describing some
of its vulnerabilities.

\paragraph{Notation.} We refer to random variables (rv) using
uppercase \( X \), to simple variables with plain lowercase \( x \)
and to vectors with bold \( \xb \). Often, we partition the input
vector \(\xb \in \mathbb{R}^D\) to the feature of interest
\(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we clarify that it implies the vector
\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
corresponding rv as \(\mathbf{X} = (X_s, \mathbf{X}_c)\). When we
refer to an instance of the training set, we use
\(\xb^i= (\xc^i, x_s^i) \). The black-box function is
\(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the FE of the \(s\)-th
feature is \(f^{\mathtt{<method>}}(x_s)\), with \(\mathtt{<method>}\)
indicating the particular method in use.\footnote{An extensive list of all
  symbols used in the paper is provided at the Appendix.}

\subsection{Feature Effect Methods And ALE Definition}
\label{sec:feat-effect-meth}

The three well-known feature effect methods are: PDP, MP and ALE. PDP
formulates the FE of the \(s\)-th attribute as an expectation over the
marginal distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
formulates it as an expectation over the conditional
\(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|x_s}[f(x_s,
\mathbf{X}_c)]\). ALE defines the local effect of the \(s\)-th feature
at point \(x_s = z\) as
\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). All the
local explanations at \(z\) are, then, weighted by the conditional
distribution \(p(\xc|x_s=z)\) and are averaged, to produce the
averaged effect \(\mu(z)\). ALE is the accumulation of the averaged
local effects:

@@

Eq.~(\ref{eq:ALE}) has specific advantages which gain particular value in cases of
correlated input features. In these cases, PDP integrates over
unrealistic instances due to the use of the marginal distribution
\(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
the combined effect of sets of features to a single feature. ALE
manages to resolve both issues, and is therefore the most trustable
method in cases of correlated features.

\subsection{Quantification Of Heterogeneous Effects.}
\label{sec:quant-heter-effects}

FE methods reply to the question \textit{what is the average (global)
  effect on the output, if the value of a specific feature is
  increased/decreased}. It comes naturally to also ask \textit{to what
  extent the local effects deviate from the global
  explanation}. Quantifying the uncertainty of the global explanation
has attracted a lot of interest. ICE and
d-ICE\citep{goldstein2015peeking} provide a set of curves that are
plot on top-of the PDP. Both methods produce one curve for each
instance of the training set;
\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) for ICE and
\(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
\xc^i)\) for d-ICE. The user then visually observes if the curves are
homogeneous and to what extent they deviate from the PDP. Some methods
try to automate the aforementioned visual exploration, by grouping
(d-)ICE plots into clusters~\citep{molnar2020model,
  herbinger2022repid, britton2019vine}. Some other approaches, like
H-Statistic\citep{friedman2008predictive}, Greenwell's interaction
index\citep{greenwell2018simple} or SHAP interaction
values\citep{lundberg2018consistent}, quantify the level of
interaction between the input features, with an interaction value. A
strong interaction index is an indicator for the existence of
heterogeneous effects.

>>>>

Recently, Machine Learning (ML) has been adopted in critical domains,
such as healthcare and finance. In these areas, we need a combination
of accurate predictions along with meaningful explanations to support
them. For this reason, there is an increased interest in Explainable
AI (XAI) to understand the decision mechanism of complex black-box
models. XAI literature distinguishes between local and global
techniques~\citep{Molnar2020interpretable}. Local methods provide
instance-level explanations, i.e.~explain the prediction for a
specific instance, whereas global methods summarize the entire model
behavior.

Global methods create a global explanation by aggregating the
instance-level explanations into a single interpretable outcome,
usually a number or a plot. When strong interactions between features
exist, the instance-level explanations may significantly deviate from
the aggregated outcome~\citep{herbinger2022repid}, a phenomenon called
\emph{aggregation bias}~\citep{mehrabi2021survey}. Therefore, global
methods should quantify not only the aggregated (averaged) effect, but
also the level of heterogeneity of instance-level explanations. The
same holds for Feature Effect (FE)~\citep{Gromping2020MAEP}, a class
of global XAI methods that isolate the impact of a single feature on
the output.\footnote{FE methods also isolate the combined effect of a
  pair of features to the output. Combinations of more than two
  features are uncommon, since they are difficult to estimate and
  visualize.} For example, a feature with zero average effect may
indicate no effect on the output or a highly positive effect in some
cases and a highly negative effect in others. The ambiguity about the
correct interpretation can be resolved, if the FE method informs about
the heterogeneity of instance-level explanations, a quantity we refer
to as the \emph{uncertainty of the effect}.

@@

We here review the existing methods for modeling the average effect
and the heterogeneity of instance-level effects and we focus on ALE
definition~\citep{apley2020visualizing} and
approximation~\citep{apley2020visualizing, gkolemis22}.

\paragraph{Notation.} We refer to random variables (rv) using
uppercase \( X \), to simple variables with plain lowercase \( x \)
and to vectors with bold \( \xb \). Often, we partition the input
vector \(\xb \in \mathbb{R}^D\) to the feature of interest
\(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we it refers the vector
\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
corresponding rv as \(\mathbf{X} = (X_s, \mathbf{X}_c)\). When we
refer to an instance of the training set, we use
\(\xb^i= (x_s^i, \xc^i) \).

% The black-box function is
% \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the FE of the \(s\)-th
% feature is \(f^{\mathtt{<method>}}(x_s)\), with \(\mathtt{<method>}\)
% indicating the particular method in use.\footnote{An extensive list of
%   all symbols used in the paper is provided at the Appendix.}

\subsection{Feature Effect Methods}
\label{sec:feat-effect-meth}

The three well-known feature effect methods are: PDP, MP and ALE. PDP
formulates the FE of the \(s\)-th feature as an expectation over the
marginal distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
formulates it as an expectation over the conditional
\(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|x_s}[f(x_s,
\mathbf{X}_c)]\). ALE defines the local effect at \(x_s = z\) as
\(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). The local
explanations are then weighted by the conditional distribution
\(p(\xc|x_s=z)\) to compute the averaged effect \(\mu(z)\). ALE is the
accumulation of the averaged local effects:

@@

Eq.~(\ref{eq:ALE}) has important advantages compared to PDP and MP in
cases of correlated features. In these cases, PDP integrates over
unrealistic instances due to the use of the marginal distribution
\(\mathbf{X}_c \), and MP computes aggregated effects, i.e., imputes
the combined effect of sets of features to a single feature. ALE
resolves both issues.

\subsection{Heterogeneity Of Instance-Level Effects.}
\label{sec:quant-heter-effects}

FE methods show what is the average (global) effect on the output, if
the value of a specific feature is increased/decreased. For completing
the explanation about a feature, we need to also know to what extent
the local effects deviate from the global explanation, i.e.~to
quantify the uncertainty of the average effect.

Heterogeneity It comes naturally to also ask \textit{to
  what extent }. Quantifying the uncertainty of the global explanation
has attracted a lot of interest. ICE and
d-ICE\citep{goldstein2015peeking} provide a set of curves that are
plot on top-of the PDP. Both methods produce one curve for each
instance of the training set;
\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) for ICE and
\(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
\xc^i)\) for d-ICE. The user then visually observes if the curves are
homogeneous and to what extent they deviate from the PDP. Some methods
try to automate the aforementioned visual exploration, by grouping
(d-)ICE plots into clusters~\citep{molnar2020model,
  herbinger2022repid, britton2019vine}. Some other approaches, like
H-Statistic\citep{friedman2008predictive}, Greenwell's interaction
index\citep{greenwell2018simple} or SHAP interaction
values\citep{lundberg2018consistent}, quantify the level of
interaction between the input features, with an interaction value. A
strong interaction index is an indicator for the existence of
heterogeneous effects.
